---
title: "HPAM 7660 Data Assignment 4"
author: "Rebecca Letsinger"
date: "March 5, 2024"
output: pdf_document
---
```{r}
    library(dplyr)
    library(knitr)
    library(ggplot2)
```

Three essential components of data visualization: 
1. Data - this includes the information and variables used and of interest. In figure 2.1, the data includes life expectancy over GDP per capita in 2007 for 147 country. The information collected under each variable represents data.
2. Geom - this describes the style of geometric object being used.This can include lines, points, and bars. In figure 2.1 this is demonstrated by the individual points on the graph, the decision to select a scatterplot, the axis and legends. 
3. Aes - this refers to the aesthetic of the chosen geom. This can include position, shape, color, and size. In figure 2.1 this can be indicated by the different color and size of the dots on the graph. 

```{r}
library(readr)
la_mort <- 
  read_csv("https://www.dropbox.com/scl/fi/fzsnhfd3lq80v2o3sag6c/la_mort.csv?rlkey=h1vyjm2b8ppgejgsg3e8evm7i&dl=1")
```

```{r}
la_mort$cancer_parish <- ifelse(la_mort$cntyrsd %in% c(5, 33, 47, 51, 71, 89, 93, 95, 121), 1, 0)
```

```{r}
table(la_mort$cancer_parish)
table(la_mort$cntyrsd[la_mort$cancer_parish == 1])
```

```{r}
la_mort$cancer39 <- ifelse(la_mort$ucr39 %in% c(5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15), 1, 0)
```

```{r}
table(la_mort$cancer39)
```

```{r}
la_mort$cancer39 <- ifelse(la_mort$ucr39 %in% c(5:15), 1, 0)
```

```{r}
table(la_mort$ucr39[la_mort$cancer113 == 1 & la_mort$cancer39 == 0])
```

```{r}
table(la_mort$ucod[la_mort$cancer113 == 1 & la_mort$cancer39 == 0])
```

```{r}
library(dplyr)
```

```{r}
parish_count <- la_mort %>%
  group_by(cntyrsd, cancer_parish, year) %>%
  summarize(cancer39 = sum(cancer39, na.rm = TRUE))
```

```{r}
summary(parish_count$cancer39)
```

```{r}
library(readr)
```

```{r}
la_pop <- 
  read_csv("https://www.dropbox.com/scl/fi/650k1obpczky6bwa19ex6/la_county_pop.csv?rlkey=0aokd9m76q7mxwus97uslsx7g&dl=1")
```

```{r}
parish_count <- parish_count %>%
  rename(county = cntyrsd)
```

```{r} 
la_joined <- parish_count %>%
  inner_join(la_pop, by = c("county", "year"))
la_joined_all <- subset(la_joined, agegrp == "all")
```

```{r}
la_joined_all$cancer_rate_total <- (la_joined_all$cancer39) / (la_joined_all$tot_pop)
```

```{r}
summary(la_joined_all$cancer_rate_total)
```

```{r}
la_joined_all$cancer_rate_total <- ((la_joined_all$cancer39) / (la_joined_all$tot_pop / 100000))
```

```{r}
parish_cancer_2019 <- subset(la_joined_all, year == 2019)
library(knitr)
kable(parish_cancer_2019[, c("county", "cancer_rate_total")])
```

```{r}
la_mort <- 
  read_csv("https://www.dropbox.com/scl/fi/fzsnhfd3lq80v2o3sag6c/la_mort.csv?rlkey=h1vyjm2b8ppgejgsg3e8evm7i&dl=1")
show_col_types = FALSE
```

```{r}
la_mort$cancer_parish <- ifelse(la_mort$cntyrsd %in% c(5, 33, 47, 51, 71, 89, 93, 95, 121), 1, 0)
```

```{r}
la_mort$cancer39 <- ifelse(la_mort$ucr39 %in% c(5:15), 1, 0)
```

```{r}
library(dplyr)
la_mort_age <- la_mort %>%
  filter(age != 9999)
la_mort_age$age <- ifelse(la_mort_age$age < 2000, la_mort_age$age - 1000, 0)
```

```{r}
age_breaks <- c(0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85, Inf)
age_labels <- c("0_4", "5_9", "10_14", "15_19", "20_24", "25_29", "30_34", "35_39", 
                "40_44", "45_49", "50_54", "55_59", "60_64", "65_69", "70_74", 
                "75_79", "80_84", "85+")
la_mort_age$agegrp <- as.character(cut(la_mort_age$age, breaks = age_breaks, labels = age_labels, right = FALSE))
```

```{r}
parish_count <- la_mort %>%
  group_by(cntyrsd, cancer_parish, year) %>%
  summarize(cancer39 = sum(cancer39, na.rm = TRUE))
```
  
```{r}
parish_count_age <- la_mort_age %>%
  group_by(cntyrsd, cancer_parish, agegrp, year) %>%
  summarize(cancer39 = sum(cancer39, na.rm = TRUE))
```

```{r}
library(readr)
```

```{r}
la_pop <- 
  read_csv("https://www.dropbox.com/scl/fi/650k1obpczky6bwa19ex6/la_county_pop.csv?rlkey=0aokd9m76q7mxwus97uslsx7g&dl=1")
```

```{r}
library(dplyr)
```
  
```{r}
la_joined <- parish_count_age %>%
  inner_join(la_pop, by = c("cntyrsd" = "county", "year", "agegrp"))
```
  
```{r}
stnrd_pop <- 
  read_csv("https://www.dropbox.com/scl/fi/xzd2o5lza237so6vamqwb/stnrd_pop.csv?rlkey=zp90au2tuq6eptvi1yiyjfzua&dl=1")
la_joined_stnrd <- la_joined %>%
  inner_join(stnrd_pop, by = "agegrp")
```
  
```{r}
la_joined_stnrd$stnrd_pop_weight <- (la_joined_stnrd$stnrd_pop) / (sum(stnrd_pop$stnrd_pop))
```

```{r}
la_joined_stnrd$cancer_rate_adj <- ((la_joined_stnrd$cancer39) / (la_joined_stnrd$tot_pop / 100000)) * la_joined_stnrd$stnrd_pop_weight
```

```{r}
parish_rates <- la_joined_stnrd %>%
  group_by(cntyrsd, cancer_parish, year) %>%
  summarize(cancer_rate_adj = sum(cancer_rate_adj, na.rm = TRUE), cancer39 = sum(cancer39), tot_pop = 
              sum(tot_pop))
parish_rates$cancer_rate_crude <- (parish_rates$cancer39) / (parish_rates$tot_pop / 100000)
```

```{r}
parish_rates$pop_weight <- (parish_rates$cancer_rate_adj) * (parish_rates$tot_pop)
cancer_alley_rates <- parish_rates %>%
  group_by(cancer_parish, year) %>%
  summarize(cancer_rate_adj_wt = sum(pop_weight) / sum(tot_pop))
```
  
```{r}
parish_rates$pop_weight <- (parish_rates$cancer_rate_adj) * (parish_rates$tot_pop)
cancer_alley_rates <- parish_rates %>%
  group_by(cancer_parish, year) %>%
  summarize(cancer_rate_adj_wt = sum(pop_weight) / sum(tot_pop))
```

```{r}
kable(cancer_alley_rates)
```

```{r}
cancer_alley <- 
  subset(cancer_alley_rates, cancer_parish == 1, select = c(cancer_rate_adj_wt, year)) %>%
  rename(cancer_alley_rate = cancer_rate_adj_wt)
no_cancer_alley <- 
  subset(cancer_alley_rates, cancer_parish == 0, select = c(cancer_rate_adj_wt, year)) %>%
  rename(no_cancer_alley_rate = cancer_rate_adj_wt)
cancer_alley_table <- cancer_alley %>%
  inner_join(no_cancer_alley, by = "year")
cancer_alley_table <- cancer_alley_table[,c("year", "cancer_alley_rate", "no_cancer_alley_rate")]
kable(cancer_alley_table)
```

```{r}
parish_rates_2019 <- subset(parish_rates, year == 2019)
ggplot(data = parish_rates_2019, aes(x = cntyrsd, y = cancer_rate_adj)) +
  geom_point() +
  labs(x = "County FIPS code", y = "Cancer Mortality Rate (2019)",
       title = "Scatterplot of Cancer Mortality Rates in 2019")
```

```{r}
filtered_data <- subset(cancer_alley_rates, year >= 2005 & year <= 2019)
```

```{r}
ggplot(data = filtered_data, aes(x = year, y = cancer_rate_adj_wt, color = factor(cancer_parish))) +
  geom_line() +
  labs(x = "Year", y = "Cancer Mortality Rate (Age-adjusted, Population Weighted)",
       color = "Parish Type", title = "Cancer Mortality Rates (2005-2019)") +
  scale_color_manual(values = c("blue", "red"), labels = c("Non-Cancer Alley", "Cancer Alley")) +
  theme_minimal()
```

```{r}
ggplot(data = filtered_data, aes(x = year, y = cancer_rate_adj_wt, color = factor(cancer_parish))) +
  geom_line() +
  labs(x = "Year", y = "Cancer Mortality Rate (Age-adjusted, Population Weighted)",
       color = "Parish Type", title = "Cancer Mortality Rates (2005-2019)") +
  scale_color_manual(values = c("blue", "red"), labels = c("Non-Cancer Alley", "Cancer Alley")) +
  scale_y_continuous(limits = c(0, 300)) + # Adjusting y-axis scale
  theme_minimal()
```

```{r}
ggplot(data = filtered_data, aes(x = year, y = cancer_rate_adj_wt, color = factor(cancer_parish))) +
  geom_line() +
  labs(x = "Year", y = "Cancer Mortality Rate (Age-adjusted, Population Weighted)",
       color = "Parish Type", title = "Cancer Mortality Rates (2005-2019)") +
  scale_color_manual(values = c("blue", "red"), labels = c("Non-Cancer Alley", "Cancer Alley")) +
  scale_y_continuous(limits = c(0, 300)) + # Adjusting y-axis scale
  scale_x_continuous(breaks = seq(2005, 2019, by = 1)) + # Ensuring all year values between 2005 and 2019 are included
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotating year labels by 45 degrees
```

```{r}
ggplot(data = filtered_data, aes(x = year, y = cancer_rate_adj_wt, color = factor(cancer_parish), linetype = factor(cancer_parish))) +
  geom_line() +
  labs(x = "Year", y = "Cancer Mortality Rate (Age-adjusted, Population Weighted)",
       color = "Parish Type", linetype = "Parish Type", title = "Cancer Mortality Rates (2005-2019)") +
  scale_color_manual(values = c("blue", "red"), labels = c("Non-Cancer Alley", "Cancer Alley")) +
  scale_linetype_discrete(name = NULL, labels = c("Rest of Louisiana", "Cancer Alley"), guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(limits = c(0, 300)) + # Adjusting y-axis scale
  scale_x_continuous(breaks = seq(2005, 2019, by = 1)) + # Ensuring all year values between 2005 and 2019 are included
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotating year labels by 45 degrees
```

```{r}
ggplot(data = filtered_data, aes(x = year, y = cancer_rate_adj_wt, color = factor(cancer_parish), linetype = factor(cancer_parish))) +
  geom_line() +
  labs(title = "Cancer Mortality Rate Comparison", y = "Cancer Deaths per 100,000 Residents", x = NULL, color = "Parish Type", linetype = "Parish Type") +
  scale_color_manual(values = c("blue", "red"), labels = c("Non-Cancer Alley", "Cancer Alley")) +
  scale_linetype_discrete(name = NULL, labels = c("Rest of Louisiana", "Cancer Alley"), guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(limits = c(0, 300)) + # Adjusting y-axis scale
  scale_x_continuous(breaks = seq(2005, 2019, by = 1)) + # Ensuring all year values between 2005 and 2019 are included
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotating year labels by 45 degrees
```

```{r}
filtered_data$cancer_parish <- factor(filtered_data$cancer_parish)
average_mortality <- filtered_data %>%
  group_by(cancer_parish) %>%
  summarise(average_rate = mean(cancer_rate_adj_wt))
ggplot(average_mortality, aes(x = cancer_parish, y = average_rate)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(x = "Parish", y = "Average Cancer Mortality Rate",
       title = "Average Cancer Mortality Rates for Cancer Alley Parishes (2005-2019)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
